Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-21 paper/paper.tex
First submitted version.

The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendixes contain more material from the full paper.

@@

GCSA is a generalization of the FM\nobreakdash-index for path graphs. Because path graphs are similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them for path graphs.

Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by the keys. We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree. The compacted trie can be considered the \emph{suffix tree} of graph $G'$.

@@

In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range. In GCSA, each node $v' \in V'$ may have multiple values (pointers), and a value may occur in multiple nodes. Determining the number of distinct values with a $\locate$ query can be slow. Hence we need additional structures to support efficient \emph{counting queries}. The problem is similar to determining the \emph{document frequency} of a pattern. Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern. We can augment an FM\nobreakdash-index with a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.

@@

The \emph{variation graphs} used in vg have strings as node labels. The average length of a label is typically around 10~bases. Each node can be traversed in both forward and reverse complement orientations, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples for performance reasons.

@@

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. The $\find$ and $\locate$ queries are several times slower than in a fast whole-genome FM\nobreakdash-index. We have to use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}, while our $\locate$ queries generate duplicates, which must be filtered out. A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.

\begin{table}[t!]
\begin{center}
\caption{Query benchmarks using an \orderk{64} index. Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
\begin{tabular}{c|cc|cc|cccc}
\hline
\textbf{Length} & \textbf{Patterns} & \textbf{Occs} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
16 & 351584 & 347453 & 1829M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 468 \textmu{}s \\
64 & 351567 & 326101 & 0.33M & 1.35M & 88.9 \textmu{}s & 0.36 \textmu{}s & 0.42 \textmu{}s & 10.3 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

For all positions $j$ with $i \le j < i+\abs{X}$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
\end{proof}

@@

Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$. By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring. We extend the substring to length $\abs{X}+k-1$ without changing its starting position. The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$. If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.

Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$. If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$. Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$. As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$. Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key. As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$. When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.

@@

Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}. Because graph $G'$ is a pruned de~Bruijn graph, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$. As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ for all nodes $v \in G'.\gvalue(v')$.

@@

We use a variant of that algorithm with GCSA2. Let $G = (V, E)$ be the input graph. We extract all paths of length $k$ (typically for $k = 16$) from graph $G$. For each path $P = v_{0} \dotsm v_{\abs{P}-1}$, we store several fields. \emph{Key} $P.\gkey$ encodes $G.\glabel(P)$ as a sequence of lexicographic ranks of \kmer{k}s. If $\abs{P}$ is not an integer multiple of $k$, the key consists of the \kmer{k} ranks for the lexicographically smallest \kmer{(\lceil \abs{P}/k \rceil \cdot k)} having $G.\glabel(P)$ as a prefix, followed by the rank of the last \kmer{k} in the largest such \kmer{(\lceil \abs{P}/k \rceil \cdot k)}. \emph{Value} $P.\gvalue$ is the start node $v_{0}$ of the path. We store the set of \emph{predecessor labels} $\set{c \in \Sigma \mid \abs{G.\gpred(v_{0}, c)} > 0}$ as $P.\gpred$. For each possible \emph{extension node} $v \in \set{v \in V \mid (v_{\abs{P}-1}, v) \in E}$, we create a new copy of the path and store the node as $P.\gext = v$.

The construction uses several supporting structures. We build an \orderk{k} \emph{de~Bruijn graph} $G_{d} = (V_{d}, E_{d})$ of the path labels and encode it as a GCSA, using the predecessor labels $P.\gpred$ for determining the edges. Let $v_{0}, \dotsc, v_{\abs{V_{d}}-1}$ be the nodes of the de~Bruijn graph in lexicographic order by the keys. We use two additional arrays: \emph{the LCP array} $\LCP[0, \abs{V_{d}}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G_{d}.\gkey(v_{i-1})$ and $G_{d}.\gkey(v_{i})$ (with $\LCP[0] = 0$), and the \emph{last character array} $L[0, \abs{V_{d}}-1]$, where $L[i] = G_{d}.\gkey(v_{i})[k-1]$. The LCP array is stored in a wavelet tree for fast \emph{range minimum queries} \cite{Gagie2012a}.

@@

The extension step transforms the current set of paths of length (up to) $k'$ and transforms it into a set of paths of length (up to) $2k'$. If $P$ is a path with $P.\gext = -1$, we use it as such. If $P.\gext = P'.\gvalue$ for paths $P$ and $P'$, we create a new path $PP'$. We set $(PP').\gkey$ according to the concatenation of the path labels, take $\gvalue$ and $\gpred$ from path $P$, and take $\gext$ from path $P'$. If we have another path $Q$ with $Q.\gkey = P.\gkey$ such that $QP'$ is a path, and if $P'.\gext = -1$, all possible \kmer{2k'} extensions of label $G.\glabel(PP')$ are also labels of paths starting from node $Q.\gvalue$, and the other way around. Hence paths $PP'$ and $QP'$ can be represented by a single node in a pruned de~Bruijn graph.

The doubling steps are followed by the \emph{merging steps}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph. It consists of a pruning step, followed by merging the paths with identical labels.} First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$. If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$. We also store the union of predecessor labels as $G'.\gpred(v') = \bigcup_{i=0}^{m-1} P_{i}.\gpred$. Then we apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.

Storing the paths and graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs. To avoid that, we store them on \emph{disk} when possible. Each chromosome forms a subgraph that is disjoint from the other chromosomes, except near the source node and the sink node. Hence we can store the paths corresponding to each chromosome in a separate file, sorted by their labels in lexicographic order. A doubling step can be done separately for each of the chromosomes. A pruning step takes a set of sorted lists paths and merges the lists. It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged. The merged path is written into the new file for that chromosome, and the original range of paths is removed from the buffer. The merging step is mostly the same as a pruning step, but the output format is different.

After storing the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$ on disk, we proceed to \emph{building the index}. Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$ we have stored, but the outdegree sequence $\bvOUT$ requires further processing. Because graph $G''$ is a path graph, there is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$. We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers. The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a $c \in \Sigma$. We can also sample the nodes for $\locate$ queries during the same scans.

@@

We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$. If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$. Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
$$
G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
$$
We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector. On the other hand, computing $G'.\LF(i)$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$. The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.

@@

The hypertext index \cite{Thachuk2013} is based on graphs $G = (V, E)$, where the label $G.\glabel(v)$ of a node $v \in V$ is a string over alphabet $\Sigma$. The labels are indexes in FM\nobreakdash-index $F$, while the reverse labels are indexed in FM\nobreakdash-index $R$. When we query for patterns $X \in \Sigma^{\ast}$ of length $\abs{X} \ge 2$, some of the matches may cross the edges. In order to find matches crossing one edge, we search for suffixes $X[i, \abs{X}-1]$ in the forward index $F$ and reverses of prefixes $X[0, i-1]$ in the reverse index $R$, for all $1 \le i \le \abs{X}-1$. For each value of $i$, we then combine the partial matches into complete matches with a two-dimensional range query $R.\find(X[0, i-1]) \times F.\find(X[i, \abs{X}-1])$ in the edge matrix $E$. Matches crossing multiple edges can be found with similar but more complicated techniques.

We often have to prune complex regions of the input graph before indexing it with GCSA. This causes false negatives: paths that exist in the input graph but not in the index. We can generalize the hypertext index to avoid the false negatives. Instead of pruning the input graph heuristically, we can create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph. We then create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph, always crossing from the reverse complement strand to the forward strand. We search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index, and combine the results $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ with a range query for all $i$.

While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately. For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph. Neither of the options is good for a reference genome. With a single copy, we create paths in the reference that do not correspond to any valid genome. With the second option, we duplicate sequences that are not only identical but also functionally the same. One solution is to switch from graphs to \emph{context-free grammars}. As long as the grammar is non-nested, we can handle it with the hypertext index. We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.

>>>>

The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendices contain more material from the full paper.

@@

Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by the keys. We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree. The compacted trie can be considered the \emph{suffix tree} of graph $G'$.

@@

In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range. In GCSA, each node $v' \in V'$ may have multiple values (pointers), and a value may occur in multiple nodes. As $\locate$ queries can be slow, we need another way to support efficient \emph{counting queries}. The problem is similar to determining the \emph{document frequency} of a pattern. Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern. We can use a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.

@@

The \emph{variation graphs} used in vg have strings as node labels. Each node can be traversed in both forward and reverse complement orientations, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples for performance reasons.

@@

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. The $\find$ and $\locate$ queries are several times slower than in a fast whole-genome FM\nobreakdash-index. We have to use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}, while our $\locate$ queries have to filter out duplicates. A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.

\begin{table}[t!]
\begin{center}
\caption{Query benchmarks using an \orderk{64} index. Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in microseconds or milliseconds.}\label{table:benchmark}
\begin{tabular}{c|cc|cc|cccc}
\hline
\textbf{Length} & \textbf{Patterns} & \textbf{Occs} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
16 & 351584 & 347453 & 1829M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 468 \textmu{}s \\
64 & 351567 & 326101 & 0.33M & 1.35M & 88.9 \textmu{}s & 0.36 \textmu{}s & 0.42 \textmu{}s & 10.3 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

For all positions $j$ with $i \le j \le i+\abs{X}-1$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
\end{proof}

@@

Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$. By Lemma~\ref{lemma:pg-keys}, $X$ is a prefix of the substring. We extend the substring to length $\abs{X}+k-1$ without changing its starting position. The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$. If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.

Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$. If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$. Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$. As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$. Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key. As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, it is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$. When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.

@@

Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}. Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$. As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ for all nodes $v \in G'.\gvalue(v')$.

@@

We use a variant of that algorithm with GCSA2. Let $G = (V, E)$ be the input graph. We extract all paths of length $k$ (typically with $k = 16$) from graph $G$. For each path $P = v_{0} \dotsm v_{\abs{P}-1}$, we store several fields. \emph{Key} $P.\gkey$ encodes $G.\glabel(P)$ as a sequence of lexicographic ranks of \kmer{k}s. If $\abs{P}$ is not an integer multiple of $k$, the key consists of the \kmer{k} ranks for the lexicographically smallest \kmer{(\lceil \abs{P}/k \rceil \cdot k)} having $G.\glabel(P)$ as a prefix, followed by the rank of the last \kmer{k} in the largest such \kmer{(\lceil \abs{P}/k \rceil \cdot k)}. \emph{Value} $P.\gvalue$ is the start node $v_{0}$ of the path. We store the set of \emph{predecessor labels} $\set{c \in \Sigma \mid \abs{G.\gpred(v_{0}, c)} > 0}$ as $P.\gpred$. For each possible \emph{extension node} $v \in \set{v \in V \mid (v_{\abs{P}-1}, v) \in E}$, we create a separate copy of the path and store the node as $P.\gext = v$.

The construction uses several supporting structures. We build an \orderk{k} \emph{de~Bruijn graph} $G_{d} = (V_{d}, E_{d})$ of the path labels and encode it as a GCSA, using the predecessor labels $P.\gpred$ for determining the edges. Let $v_{0}, \dotsc, v_{\abs{V_{d}}-1}$ be the nodes of the de~Bruijn graph in lexicographic order by the keys. We use two additional arrays: the \emph{LCP array} $\LCP[0, \abs{V_{d}}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G_{d}.\gkey(v_{i-1})$ and $G_{d}.\gkey(v_{i})$ (with $\LCP[0] = 0$), and the \emph{last character array} $L[0, \abs{V_{d}}-1]$, where $L[i] = G_{d}.\gkey(v_{i})[k-1]$. The LCP array is stored as a wavelet tree for fast \emph{range minimum queries} \cite{Gagie2012a}.

@@

The extension step transforms the current set of paths of length (up to) $k'$ into a set of paths of length (up to) $2k'$. If $P$ is a path with $P.\gext = -1$, we use it as such. If $P.\gext = P'.\gvalue$ for paths $P$ and $P'$, we create a new path $PP'$. We set $(PP').\gkey$ according to the concatenation of the path labels, take $\gvalue$ and $\gpred$ from path $P$, and take $\gext$ from path $P'$. If we have another path $Q$ with $Q.\gkey = P.\gkey$ such that $QP'$ is a path, and if $P'.\gext = -1$, all possible \kmer{2k'} extensions of label $G.\glabel(PP')$ are also labels of paths starting from node $Q.\gvalue$, and the other way around. Hence paths $PP'$ and $QP'$ can be represented by a single node in a pruned de~Bruijn graph.

The doubling steps are followed by the \emph{merging step}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph. It consists of a pruning step, followed by merging the paths with identical labels.} First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$. If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$. We also store the union of predecessor labels as $G'.\gpred(v') = \bigcup_{i=0}^{m-1} P_{i}.\gpred$. Then we apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.

Storing the paths and graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs. To avoid that, we store them on \emph{disk} when possible. Each chromosome forms a subgraph that is disjoint from the other chromosomes, except near the source node and the sink node. Hence we can store the paths corresponding to each chromosome in a separate file, sorted by their labels in lexicographic order. An extension step can be done separately for each of the chromosomes. A pruning step takes a set of sorted lists of paths and merges the lists. It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged. The merged path is written into the new file for that chromosome, and the original range of paths is removed from the buffer. The merging step is mostly the same as a pruning step, but the output format is different.

After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we proceed to \emph{building the index}. Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$ we have stored, but the outdegree sequence $\bvOUT$ requires further processing. Because graph $G''$ is a path graph, there is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$. We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers. The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a $c \in \Sigma$. We can also sample the nodes for $\locate$ queries during the same scans.

@@

We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$. If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$. Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
$$
G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
$$
We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector. On the other hand, computing $G'.\LF(i)$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$. The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.

@@

The hypertext index \cite{Thachuk2013} is based on graphs $G = (V, E)$, where the label $G.\glabel(v)$ of a node $v \in V$ is a string over alphabet $\Sigma$. The labels are indexes in FM\nobreakdash-index $F$, while the reverse labels are indexed in FM\nobreakdash-index $R$. When we search for a pattern $X \in \Sigma^{\ast}$ of length $\abs{X} \ge 2$, some of the matches may cross edges. In order to find matches crossing one edge, we search for suffixes $X[i, \abs{X}-1]$ in the forward index $F$ and the reverses of prefixes $X[0, i-1]$ in the reverse index $R$, for all $1 \le i \le \abs{X}-1$. For each value of $i$, we combine the partial matches into complete matches with a two-dimensional range query in the edge matrix $E$, using the lexicographic ranges for the reverse of $X[0, i-1]$ and for $X[i, \abs{X}-1]$ as the query ranges.

We often have to prune complex regions of the input graph before indexing it with GCSA. This causes false negatives: paths that exist in the input graph but not in the index. We can avoid the false negatives with a generalization of the hypertext index. Instead of pruning the input graph heuristically, we create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph. We then create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph, always crossing from the reverse complement strand to the forward strand. We search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index, and combine the results $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ with a range query.

While graphs are a natural formalism for representing genetic variation, they cannot adequately represent certain types of \emph{rearrangements}. For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph. Neither option is particularly good in a reference genome. With a single copy, we create paths in the reference that do not correspond to any valid genome. With the second option, we duplicate sequences that are not only identical but also functionally the same. One solution is to switch from graphs to \emph{context-free grammars}. As long as the grammar is non-nested, we can handle it with the hypertext index. We build a hypertext index for a high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
